disaggregation: An \proglang{R} Package for Bayesian Spatial Disaggregation Modelling

### 2019-12-10 article.Rnw
Easier to read citations

Methods for estimating high-resolution risk maps from aggregated response data over large spatial regions are 
becoming increasingly sought after in disease risk mapping~\cite{li2012log, diggle2013spatial, wilson2017pointless}, especially in malaria mapping~\cite{sturrock2014fine, sturrock2016mapping}. 
Disaggregation regression, first applied in species distribution modelling in ecology~\cite{keil2013downscaling, barwell2014can} has now become an important method in disease risk mapping \cite{falciparum2019, vivax2019}. 
The aggregation of response data over large heterogenous regions is problematic for making fine-scale predictions, as relationships learned between variables at one spatial scale may not hold at other scales \cite{wakefield2006health}. However, by using fine-scale information from related covariates we can inform the hetrogeneity of the response variable of interest within the aggregated area.

Disaggregation modelling is unorthodox as the predictions are at a different scale to the response data, i.e. the number of rows in the covariate data is different to that of the response data. The spatial modelling software package, \pkg{INLA}~\cite{rue2009approximate} (integrated nested Laplace approximation), has been shown to very useful in a wide variety of circumstances, however it is not flexible enough for the unorthodox nature of the disaggregation problem except in the special case of the linear link function \cite{wilson2017pointless}. Disaggregation models can be implemented in \pkg{TMB}~\cite{tmb2016} (template model builder) with a lot of flexibility, however the data manipulation required to format the model objects and construct the model definition in C++ is non-trivial. The \pkg{disaggregation} package allows this process to be streamlined, to make it easy for the user to run disaggregation models at the expense of some flexibility.

@@

For each of the model parameters and hyperparameters we specify priors. The regression parameters and intercept are given Gaussian priors, where the default priors are $\beta_0 \sim N(0, 2)$ and $\beta_i \sim N(0, 0.4)$. A penalised complexity prior is placed on the scale and range parameters of the random field \cite{fuglstad2018constructing} such that
\begin{align}
    P(\rho < \rho_{min}) &= \rho_{prob} \\
    P(\sigma > \sigma_{max}) &= \sigma_{prob}
\end{align}
where the values $\rho_{min}, \,\rho_{prob}, \,\sigma_{max}, \,\sigma_{prob}$ are set by the user. This prior shrinks the field towards a base model with zero variance and infinite range, in other words regularising towards a flatter field with smaller magnitude. The polygon-specific effects $u_1,...,u_N$ have Gaussian priors centered at 0 with standard deviation $\sigma_{u}$ (where the precision $\tau_{u} = 1/\sigma_{u}^2$). A penalised complexity prior \cite{simpson2017penalising} such that 
\begin{equation}
P(\sigma_{u} > \sigma_{u, max}) = \sigma_{u, prob}
\end{equation}
which shrinks towards a base model of no polygon-specific effect. The $\sigma_{u, max}$ and $\sigma_{u, prob}$ are set by the user.

@@

The \pkg{disaggregation} package is built on template model builder (\pkg{TMB})~\cite{tmb2016}, which is a tool for flexibly building complex models based on C++. \pkg{TMB} combines the packages CppAD~\cite{bell2012cppad}, for automatic differentiation in C++, Eigen~\cite{guennebaud2010eigen}, a C++ library for linear algebra, and CHOLMOD~\cite{chen2008algorithm}, a package for efficient computation of sparse matrices. The use of these packages allows an efficient implementation of the automatic Laplace approximation~\cite{skaug2006automatic} with exact derivatives which gives an approximation to the Bayesian posterior. \pkg{TMB} calculates first and second order derivatives of the objective function using automatic differentiation~\cite{griewank2008evaluating}.

@@

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS (no-u-turn sampler) algorithm \cite{hoffman2014no} using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{fit_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 94 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations, doubling the number of iterations each time until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. In contrast, fitting the model using the Laplace approximation via \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}.

>>>>

Methods for estimating high-resolution risk maps from aggregated response data over large spatial regions are 
becoming increasingly sought after in disease risk mapping~\citep{li2012log, diggle2013spatial, wilson2017pointless}, especially in malaria mapping~\citep{sturrock2014fine, sturrock2016mapping}. 
Disaggregation regression, first applied in species distribution modelling in ecology~\citep{keil2013downscaling, barwell2014can} has now become an important method in disease risk mapping \citep{falciparum2019, vivax2019}. 
The aggregation of response data over large heterogenous regions is problematic for making fine-scale predictions, as relationships learned between variables at one spatial scale may not hold at other scales \citep{wakefield2006health}. However, by using fine-scale information from related covariates we can inform the hetrogeneity of the response variable of interest within the aggregated area.

Disaggregation modelling is unorthodox as the predictions are at a different scale to the response data, i.e. the number of rows in the covariate data is different to that of the response data. The spatial modelling software package, \pkg{INLA}~\citep{rue2009approximate} (integrated nested Laplace approximation), has been shown to very useful in a wide variety of circumstances, however it is not flexible enough for the unorthodox nature of the disaggregation problem except in the special case of the linear link function \citep{wilson2017pointless}. Disaggregation models can be implemented in \pkg{TMB}~\citep{tmb2016} (template model builder) with a lot of flexibility, however the data manipulation required to format the model objects and construct the model definition in C++ is non-trivial. The \pkg{disaggregation} package allows this process to be streamlined, to make it easy for the user to run disaggregation models at the expense of some flexibility.

@@

For each of the model parameters and hyperparameters we specify priors. The regression parameters and intercept are given Gaussian priors, where the default priors are $\beta_0 \sim N(0, 2)$ and $\beta_i \sim N(0, 0.4)$. A penalised complexity prior is placed on the scale and range parameters of the random field \citep{fuglstad2018constructing} such that
\begin{align}
    P(\rho < \rho_{min}) &= \rho_{prob} \\
    P(\sigma > \sigma_{max}) &= \sigma_{prob}
\end{align}
where the values $\rho_{min}, \,\rho_{prob}, \,\sigma_{max}, \,\sigma_{prob}$ are set by the user. This prior shrinks the field towards a base model with zero variance and infinite range, in other words regularising towards a flatter field with smaller magnitude. The polygon-specific effects $u_1,...,u_N$ have Gaussian priors centered at 0 with standard deviation $\sigma_{u}$ (where the precision $\tau_{u} = 1/\sigma_{u}^2$). A penalised complexity prior \citep{simpson2017penalising} such that 
\begin{equation}
P(\sigma_{u} > \sigma_{u, max}) = \sigma_{u, prob}
\end{equation}
which shrinks towards a base model of no polygon-specific effect. The $\sigma_{u, max}$ and $\sigma_{u, prob}$ are set by the user.

@@

The \pkg{disaggregation} package is built on template model builder (\pkg{TMB})~\cite{tmb2016}, which is a tool for flexibly building complex models based on C++. \pkg{TMB} combines the packages CppAD~\citep{bell2012cppad}, for automatic differentiation in C++, Eigen~\citep{guennebaud2010eigen}, a C++ library for linear algebra, and CHOLMOD~\citep{chen2008algorithm}, a package for efficient computation of sparse matrices. The use of these packages allows an efficient implementation of the automatic Laplace approximation~\citep{skaug2006automatic} with exact derivatives which gives an approximation to the Bayesian posterior. \pkg{TMB} calculates first and second order derivatives of the objective function using automatic differentiation~\citep{griewank2008evaluating}.

@@

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS (no-u-turn sampler) algorithm \citep{hoffman2014no} using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{fit_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 94 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations, doubling the number of iterations each time until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. In contrast, fitting the model using the Laplace approximation via \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}.
